A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-10 text/manuscript.Rmd

Recent advances in technology have made the study of cancer possible at
unprecedented levels of detail and scale. Array and next-generation sequencing
technologies have made it possible to assay germline and somatic mutations,
expression profiles, DNA methylation, and copy number variations. To take
advantage of these tools several large consortia, including the cancer genome
atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large
numbers of tumors and collected data from multiple assays that could be analyzed
together, with the goal of integration and increased understanding of the
mechanisms of cancer. These efforts have focused on recognizing and quantifying
the importance of genes in cancer etiology as the result of selection processes,
or classification of cancers relative to common patterns of aberration in
multiple assays.

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
when genes are mutated at a greater rate than expected due to chance. Based in
part on studies like these, it has been estimated that on the order of three
mutations in key oncogenes and/or tumor suppressors would be sufficient to
transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].

Other studies aim at providing a comprehensive view of mutations, gene
expression and genomic signatures, with the express goal of understanding the
common themes of all cancer independent of tissue of origin. Such a view of
cancer as a malady affecting common cellular processes has long been a goal of
the field and is perhaps best characterized in an essay by @weinberg2000 (see also
Discussion). The first of these studies, (the
“Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
using integrative analysis with co-equal weighting of gene expression,
methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though a significant
proportion of tumors ($10\%$) could be reclassified independent of
tissue-of-origin. In a second study involving 33 cancer types and a much greater
number of tumors using a newer algorithm (iCluster)
[@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
could be further subdivided into organ specific groups, including
pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].

More recently, attempts have been made to consider pan-cancer mutations and
other data outside the context of enrichment at the gene level. Horn _et al._
[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
sparseness of mutations, allowing the potential importance of a gene to be
influenced by the aggregate of positive selection in its nearest network
neighbors. This method is powerful especially when applied to a small number (on
the order of hundreds) of tumors. However, the larger problem of cancer etiology
remains: what mutations are required, in what order, and are there specific
combinations of delinquent cellular processes sufficient to cause cancer? If
there is a common logic to cancer, how is it to be detected underneath the
overwhelming cell-of-origin signal?

In this study, we attempt to address these questions by focusing on discrete
molecular pathways as the unit of disruption in cancer. Our approach strips all
cell-type-specific information from the mutation data and generalizing
gene-level information to non-tissue specific cell-biological pathway
phenotypes. We use this rubric to evaluate all cancers and show that,
surprisingly, tumors may have distinct phenotypes completely independent of
their tissues of origin.

>>>>

```{r child='introduction.Rmd'}
```
